Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers

### 2025-02-04 endofquarterpaper.qmd
seriation now plots with colour for country of site

```{r}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"
## seriation
library(ggpubr)
artifact_matrix <- as.matrix(artifact_data %>% 
 mutate(across(everything(), ~replace(., . > 1, 1))))
rownames(artifact_matrix) <- rownames(artifact_data)
mplot <- function(x,...){
  d <- data.frame(x = rep(1:ncol(x), each = nrow(x)),
                  y = rep(nrow(x):1, ncol(x)),
                  z = as.integer(x))
  ggplot(d, aes(x, y, fill = factor(z))) +
    geom_tile(col = "grey50", linewidth = 0.5) +
    scale_fill_manual(values = c("0" = "white", "1" = "black")) +
    coord_equal(expand = FALSE) +
    scale_x_continuous(breaks = 1:ncol(x), 
                       labels = colnames(x),
                       position = "top") +
    scale_y_continuous(breaks = 1:nrow(x), 
                       labels = rev(rownames(x))) +
    theme(axis.text = element_text(size = rel(0.5)),
          axis.ticks = element_blank(),
          legend.position = "none",
          plot.title = element_text(hjust = 0.25)) +
    labs(x = "Sign Type",
         y = "Site Name",
         fill = "Count") +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)}
increase_focus <- function(x){
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[order(mcp), ]
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE) |> unlist()
  x[, order(mrp)]}
concentrate <- function(x){
  old <- x
  not_identical <- TRUE
  while(not_identical){
    new <- increase_focus(old)
    not_identical <- !identical(old, new)
    old <- new}
  new}
artifact_matrix |> concentrate() |> mplot()

@@

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by group"
graph_data <- artifact_data %>% 
  mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
  concentrate() %>% 
  rownames_to_column("site_name") %>% 
  left_join(group_df)
sign_type_data <- graph_data %>% 
  pivot_longer(cols= notch:hashtag,
               names_to = "sign_type",
               values_to = "count") %>% 
  dplyr::filter(count != 0) %>% 
  mutate(group = parse_number(group))
x_order <- graph_data %>% 
  dplyr::select(-site_name,
         -group) %>% 
  colnames() 
ggplot(sign_type_data) +
  aes(x = sign_type) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = x_order) +
  scale_x_discrete(limits = c(x_order)) +
  scale_y_continuous(labels = NULL) +
  facet_wrap(~group, nrow = 11)

>>>>

```{r}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"
## seriation
library(ggpubr)
artifact_matrix <- as.matrix(artifact_data %>% 
 mutate(across(everything(), ~replace(., . > 1, 1))))
rownames(artifact_matrix) <- rownames(artifact_data)

# Load required libraries
library(ggplot2)

# Define the mplot function
mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
  # Create the data frame
  d <- data.frame(
    x = rep(1:ncol(x), each = nrow(x)),
    y = rep(nrow(x):1, ncol(x)),
    z = as.integer(x)
  )

  # Add country colors to the data frame if provided
  if (!is.null(country_colors)) {
    d$country_color <- rep(country_colors, ncol(x))
  }

  # Use custom row/column labels if provided, otherwise use defaults
  if (is.null(col_labels)) col_labels <- colnames(x)
  if (is.null(row_labels)) row_labels <- rev(rownames(x))

  # Plot
  p <- ggplot(d, aes(x, y)) +
    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
    scale_fill_identity() +  # Use the fill values directly
    coord_equal(expand = FALSE) +
    scale_x_continuous(
      breaks = 1:ncol(x),
      labels = col_labels,
      position = "top"
    ) +
    scale_y_continuous(
      breaks = 1:nrow(x),
      labels = row_labels
    ) +
    theme(
      axis.text = element_text(size = rel(0.5)),
      axis.ticks = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0.25)
    ) +
    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)

  p
}

# Define the increase_focus function
increase_focus <- function(x, country = NULL) {
  # Reorder rows by the mean position of 1s in each row
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  row_order <- order(mcp, na.last = TRUE)
  x <- x[row_order, ]

  # Reorder the country vector if provided
  if (!is.null(country)) {
    country <- country[row_order]
  }

  # Reorder columns by the mean position of 1s in each column
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[, order(mrp, na.last = TRUE)]

  # Return the reordered matrix and country vector
  list(matrix = x, country = country)
}

# Define the concentrate function
concentrate <- function(x, country = NULL, max_iter = 100) {
  old <- x
  not_identical <- TRUE
  iter <- 0

  while (not_identical && iter < max_iter) {
    result <- increase_focus(old, country)
    new <- result$matrix
    country <- result$country
    not_identical <- !identical(old, new)
    old <- new
    iter <- iter + 1
  }

  if (iter == max_iter) {
    warning("Reached maximum iterations without convergence.")
  }

  list(matrix = new, country = country)
}

# get country names, for example
x_df <- 
  artifact_data %>% 
  rownames_to_column("site_name") %>% 
  left_join(signbase_full_clean %>% 
              distinct(site_name, .keep_all = TRUE) %>% 
              select(site_name,
                     country),
            by = "site_name")

countries <- x_df$country

# Generate a named vector of colors, one for each unique country
unique_countries <- unique(countries)
color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")   # Generate distinct colors
row_colors <- setNames(color_palette, unique_countries) 
countries_colors <- row_colors[countries]

# Concentrate the matrix and reorder the country vector
result <- concentrate(artifact_matrix, country = countries)
concentrated_matrix <- result$matrix
reordered_countries <- result$country

# Map countries to colors
country_colors <- row_colors[reordered_countries]

# Plot the concentrated matrix with country colors
mplot(
  concentrated_matrix,
  country_colors = country_colors
)
```

@@

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by group"
graph_data <- artifact_data %>% 
  mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
 # concentrate() %>% 
  rownames_to_column("site_name") %>% 
  left_join(group_df)

sign_type_data <- graph_data %>% 
  pivot_longer(cols= notch:hashtag,
               names_to = "sign_type",
               values_to = "count") %>% 
  dplyr::filter(count != 0) %>% 
  mutate(group = parse_number(group))

x_order <- graph_data %>% 
  dplyr::select(-site_name,
         -group) %>% 
  colnames()

ggplot(sign_type_data) +
  aes(x = sign_type) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = x_order) +
  scale_x_discrete(limits = c(x_order)) +
  scale_y_continuous(labels = NULL) +
  facet_wrap(~group, nrow = 11)
